---
title: "Storm Data Analysis"
author: "Suzannah Herron"
date: "2/4/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis

This data comes from the US National Oceanic and Atmospheric Administration's (NOAA) storm database. There was a lot of cleaning that needed to be done, especially regarding the names of the event types. \
The goal is to look at:\
  1. Which weather event types are the deadliest (first plot)\
  2. Which weather event types have the greatest economic impact (second plot)\
I also decided to find which states were the safest and most dangerous to live in (third plot).




# Data Processing

Checking to see if dataset exists already. If not, then download and unzip.
```{r cache=TRUE}
file_name <- "storms.csv.bz2"
if (!file.exists(file_name)) {
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(url, file_name, method = "curl")
}
if (!file.exists(file_name)) {
  unzip(file_name)
}
storms <- read.csv(file_name)
str(storms)
```
Format `BGN_DATE` to be a date then add a column with the year to use for factoring later.
```{r message=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
```
```{r}
storms$BGN_DATE <- mdy_hms(storms$BGN_DATE)
storms$YEAR <- year(storms$BGN_DATE)
# str(storms)
```

Cutting down the dataset to relevant columns/variables for analysis for bodily, property, and crop damage.

### Variables

**STATE**: State abbreviation (I am keeping this in case I want to try mapping)\
**EVTYPE**: Type of event \
**FATALITIES**: Number of fatalities caused by event\
**INJURIES**: Number of injuries caused by event\
**PROPDMG**: USD amount of property damage\
**PROPDMGEXP**: Multiplying factor {exponent} for PROPDMG\
**CROPDMG**: USD amount of crop damage\
**CROPDMGEXP**: Multiplying factor (exponent) for CROPDMG\
**YEAR**: Starting year of event extracted from `storms$BGN_DATE`\
```{r}
cols <- c( "STATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", "YEAR")
storms1 <- storms[ , cols]
# str(storms1)
```

Since we are looking at the different effects of each event type, a creation of a dataframe grouped by event types is necessary.\
Checking the types of events:
```{r}
types <- unique(storms1$EVTYPE)
str(types)
```
Some of the event types are all upper case and some are a combo of upper and lower case. Changed all to lowercase and the number of events went from 985 to 898
```{r}
storms2 <- storms1
storms2$EVTYPE <- tolower(storms2$EVTYPE) 
types <- unique(storms2$EVTYPE)
str(types)
```
There are several event names with whitespace in the beginning so that was trimmed and the number of events is 890
```{r}
storms2$EVTYPE <- trimws(storms2$EVTYPE, which = "both")
types <- unique(storms2$EVTYPE)
str(types)
```
The next step would be to group the event types into larger categories to make it easier to analyze.\
Categories will be:\
*tornado, wind, hail, rain, flood, hurricane, lightning, cold (snow, ice, blizzards, winter storms), fire*\
Afterwards, I will check the events categorized as "other" and reassess.
```{r}
storms3 <- storms2
storms3$CAT <- "other"
storms3$CAT[grep("tornado", storms3$EVTYPE)] <- "tornado"
storms3$CAT[grep("wind", storms3$EVTYPE)] <- "wind"
storms3$CAT[grep("hail", storms3$EVTYPE)] <- "hail"
storms3$CAT[grep("rain", storms3$EVTYPE)] <- "rain"
storms3$CAT[grep("flood", storms3$EVTYPE)] <- "flood"
storms3$CAT[grep("hurricane", storms3$EVTYPE)] <- "hurricane"
storms3$CAT[grep("lightning", storms3$EVTYPE)] <- "thunderstorm"
storms3$CAT[grep("snow", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("ice", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("blizzard", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("winter", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("fire", storms3$EVTYPE)] <- "fire"
sort(table(storms3$CAT))
```
Take a look what is still in the "other" category:
```{r}
other <- storms3
other <- storms3[storms3$CAT == "other", ]
other <- other[ , c(2,10)]
types_other <- unique(other$EVTYPE)
head(types_other)
```
Categorize more and changing "fire" to "heat" and changing "lightning" to "thunderstorm". Adding categories "volcano" and "dry". (I changed some of the code in chunk 9.)
```{r}
storms3$CAT[grep("cold", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("record low", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("fire", storms3$EVTYPE)] <- "heat"
storms3$CAT[grep("heat", storms3$EVTYPE)] <- "heat"
storms3$CAT[grep("hot", storms3$EVTYPE)] <- "heat"
storms3$CAT[grep("warm", storms3$EVTYPE)] <- "heat"
storms3$CAT[grep("record high", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("icy", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("freez", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("sleet", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("showers", storms3$EVTYPE)] <- "rain"
storms3$CAT[grep("precipitation", storms3$EVTYPE)] <- "cold"
storms3$CAT[grep("thunderstorm", storms3$EVTYPE)] <- "thunderstorm"
storms3$CAT[grep("microburst", storms3$EVTYPE)] <- "thunderstorm"
storms3$CAT[grep("tstm", storms3$EVTYPE)] <- "thunderstorm"
storms3$CAT[grep("volcan", storms3$EVTYPE)] <- "volcano"
storms3$CAT[grep("dry", storms3$EVTYPE)] <- "dry"
storms3$CAT[grep("driest", storms3$EVTYPE)] <- "dry"
storms3$CAT[grep("drought", storms3$EVTYPE)] <- "dry"
```

```{r}
other <- storms3
other <- storms3[storms3$CAT == "other", ]
other <- other[ , c(2,10)]
types_other <- unique(other$EVTYPE)
length(types_other)
length(unique(storms3$CAT))
```
Now I have 12 categories with 222 event types categorized as "other". Just as a side note, looking through the event types, there are a lot of typos in this data.

## Splitting Data

Create two different dataframes with injuries/fatalities and financial impact.
```{r}
fatal_inj <- storms3[ , c(1:4, 9, 10)]
head(fatal_inj)
```
```{r}
prop_crop <- storms3[ , c(1, 2, 5:10)]
head(prop_crop)
```



# Analysis


## Which types of events are most harmful to population health?

The first step is to find the totals for injuries and fatalities for each event type per year.
```{r}
sum_fatal <- aggregate(FATALITIES ~ CAT + YEAR, fatal_inj, sum)
sum_inj <- aggregate(INJURIES ~ CAT + YEAR, fatal_inj, sum)
sum_fi <- cbind(sum_fatal, sum_inj$INJURIES)
colnames(sum_fi) <- c("cat", "year", "fatalities", "injuries")
head(sum_fi)
```
Add a row with the sum of total health impact (fatalities + injuries)
```{r}
total_fi <- sum_fi
total_fi$total <- total_fi$fatalities + total_fi$injuries
head(total_fi)
```


Then create a plot of total human impact per event each year.

```{r}
# png(filename= "fi.png", height = 600, width = 600)
ggplot(data = total_fi) +
  geom_line(mapping = aes(x = year, y = total)) +
  facet_wrap(~cat) +
  labs(title = "Total Fatalities and Injuries Per Weather Event",
       x = "Year",
       y = "Total Fatalities/Injuries")
# dev.off()
```

You can clearly see that tornados are the most harmful to humans and you can confirm it by ordering the totals:
```{r}
events_fi <- aggregate(total ~ cat, total_fi, sum)
events_fi <- events_fi[order(-events_fi$total), ]
events_fi
```

It is also interesting to see when certain categories were starting to be recorded.



## Which types of events have the greatest economic consequences?


First, let's take a look at `prop_crop$CROPDMGEXP` and `prop_crop$PROPDMGEXP` to see what's in them and if they need formatting:
```{r}
unique(prop_crop$CROPDMGEXP)
unique(prop_crop$PROPDMGEXP)
```
As expected, they need some help. 


```{r}
prop_crop2 <- prop_crop
prop_crop2$pexp <- prop_crop2$PROPDMGEXP
prop_crop2$pexp[grep("[Kk]|3", prop_crop2$PROPDMGEXP)] <- 10^3
prop_crop2$pexp[grep("[Mm]|6", prop_crop2$PROPDMGEXP)] <- 10^6
prop_crop2$pexp[grep("[Bb]|9", prop_crop2$PROPDMGEXP)] <- 10^9
prop_crop2$pexp[grep("[Hh]|2", prop_crop2$PROPDMGEXP)] <- 10^2
prop_crop2$pexp[grep("8", prop_crop2$PROPDMGEXP)] <- 10^8
prop_crop2$pexp[grep("5", prop_crop2$PROPDMGEXP)] <- 10^5
prop_crop2$pexp[grep("4", prop_crop2$PROPDMGEXP)] <- 10^4
prop_crop2$pexp[grep("7", prop_crop2$PROPDMGEXP)] <- 10^7
prop_crop2$pexp[grep("0", prop_crop2$PROPDMGEXP)] <- 1
prop_crop2$pexp[grep("1", prop_crop2$PROPDMGEXP)] <- 10
prop_crop2$pexp[grep("-", fixed = TRUE, prop_crop2$PROPDMGEXP)] <- 1
prop_crop2$pexp[grep("?", fixed = TRUE, prop_crop2$PROPDMGEXP)] <- 1
prop_crop2$pexp[grep("+", fixed = TRUE, prop_crop2$PROPDMGEXP)] <- 1
prop_crop2$pexp <- as.numeric(prop_crop2$pexp)
```
```{r}
prop_crop2$cexp <- prop_crop2$CROPDMGEXP
prop_crop2$cexp[grep("[Kk]", prop_crop2$CROPDMGEXP)] <- 10^3
prop_crop2$cexp[grep("[Mm]", prop_crop2$CROPDMGEXP)] <- 10^6
prop_crop2$cexp[grep("[Bb]", prop_crop2$CROPDMGEXP)] <- 10^9
prop_crop2$cexp[grep("0", prop_crop2$CROPDMGEXP)] <- 1
prop_crop2$cexp[grep("?", fixed = TRUE, prop_crop2$CROPDMGEXP)] <- 1
prop_crop2$cexp[grep("2", prop_crop2$CROPDMGEXP)] <- 10^2
unique(prop_crop2$cexp)
prop_crop2$cexp <- as.numeric(prop_crop2$cexp)

```

Take out the rows that have `NA`s for both of the created exponential columns
```{r}
dmg <- prop_crop2[(!(is.na(prop_crop2$pexp) & is.na(prop_crop2$cexp))), ]
```

Next it's time to calculate the totals of crop and property damage by multiplying the `DMG` columns by the `DMGEXP` columns. Then change the `NA`s to 0 so they columns can be added.
```{r}
pc4 <- dmg
pc4$prop_total <- pc4$PROPDMG * pc4$pexp
pc4$crop_total <- pc4$CROPDMG * pc4$cexp
pc4$crop_total[is.na(pc4$crop_total)] <- 0
pc4$prop_total[is.na(pc4$prop_total)] <- 0
pc4$total <- pc4$prop_total + pc4$crop_total
pc4 <- pc4[ , c(1,2,7,8,13)]
str(pc4)
```

Find the sum per category and divide the total by 1M so it will be more readable in the graph.
```{r}
event_dmg <- aggregate(total ~ CAT + YEAR, pc4, sum)
event_dmg$total <- event_dmg$total/10^6
str(event_dmg)
```
Let's graph it 
```{r}
# png(filename= "cost.png", height = 600, width = 600)
ggplot(data = event_dmg) +
  geom_line(mapping = aes(x = YEAR, y = total)) +
  facet_wrap(~CAT) +
  labs(title = "Economic Impact Per Weather Event",
       x = "Year",
       y = "Total Cost by $1M")
# dev.off()
```

```{r}
total_dmg <- aggregate(total ~ CAT, event_dmg, sum)
total_dmg <- total_dmg[order(-total_dmg$total), ]
total_dmg
```
Here you can see that flooding has been the most costly weather event\

## Where should you live?


I thought it would be interesting to see which states are the best and worst to live in regarding weather events. Since tornadoes and floods are the most dangerous, I have split the data to contain only those categories.
```{r}
states <- prop_crop
states <- states[states$CAT == "tornado" | states$CAT == "flood", c(1, 7, 8)]
head(states)
states[ , 1] <- as.factor(states$STATE)
states[ , 3] <- as.factor(states$CAT)
summary(states)
```

```{r warning=FALSE}
ggplot(data = states) +
  geom_histogram(mapping = aes(x = STATE, fill = CAT), stat = "count") +
  labs(title = "Total of Flood and Tornado Events Per State",
       x = "State",
       y = "Total Number of Events",
       fill = "Event") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5))

```
It looks like Texas is the most dangerous state to live in. We can confirm that by ordering the number of events for each event type.

```{r}
flood <- states[states$CAT == "flood", ]
tornado <- states[states$CAT == "tornado", ]
```
```{r}
t <- tornado %>%
  count(STATE)
t <- t[order(t$n), ]
head(t, 3)
```
```{r}
tail(t, 3)
```
```{r}
f <- flood %>%
  count(STATE)
f <- f[order(f$n), ]
head(f, 3)
```
```{r}
tail(f, 3)
```



# Results

Tornadoes result in the most fatalities and injuries while flooding has the greatest economic impact. Ergo, stay away from areas with tornadoes and floods.\
The safest states/districts/territories from tornadoes are:\
  1. Washington DC\
  2. Alaska\
  3. Rhode Island\
\
The most dangerous states/districts/territories for tornadoes are:\
  1. Texas\
  2. Kansas\
  3. Oklahoma\
\
The safest states/districts/territories from floods are:\
  1. Guam\
  2. American Samoa\
  3. Virgin Islands\
\
The most dangerous states/districts/territories for tornadoes are:\
  1. Texas\
  2. Missouri\
  3. Kentucky\

We learned that Texas is a very dangerous state to live in and Rhode Island is a nice and safe little state.









